---
title: "CS601C Computational Statistics Midterm Project"
author: "Vijaysingh Puwar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
library(dplyr)
library(ggplot2)
library(corrplot)
library(MASS)
library(car)       # For multicollinearity check (VIF)
library(caret)     # For model training & cross-validation
library(knitr)
```

---

# **1. College Dataset Analysis**

## **1a. Load & Clean Data**
```{r}
college_path <- "C:/Users/vp32780n/Downloads/College.csv"
if (!file.exists(college_path)) stop(paste("Error: College.csv not found at", college_path))

college <- read.csv(college_path, stringsAsFactors = TRUE)
set.seed(123)
college <- college %>% sample_frac(0.9)  # Remove 10% of rows randomly
college <- college %>% mutate(Elite = ifelse(Top10perc > 50, "Yes", "No"))
str(college)
```

## **1b. Exploratory Data Analysis (EDA)**

### **Summary Statistics**
```{r}
summary(college %>% select_if(is.numeric))
```

### **Scatterplot Matrix**
```{r}
pairs(college[, 2:11], main = "Scatterplot Matrix of First 10 Quantitative Columns")
```

### **Boxplots: Tuition Comparison**
```{r}
boxplot(Outstate ~ Private, data = college, col = c("lightblue", "pink"),
        main = "Out-of-State Tuition by Private/Public", xlab = "Private", ylab = "Outstate Tuition")
boxplot(Outstate ~ Elite, data = college, col = c("green", "orange"),
        main = "Out-of-State Tuition by Elite", xlab = "Elite", ylab = "Outstate Tuition")
```

### **Histograms**
```{r}
par(mfrow=c(2,2))
hist(college$Apps, breaks=20, main="Applications", col="blue")
hist(college$Enroll, breaks=20, main="Enrollment", col="red")
hist(college$Outstate, breaks=20, main="Out-of-state Tuition", col="green")
hist(college$Grad.Rate, breaks=20, main="Graduation Rate", col="purple")
```

### **Correlation Heatmap**
```{r}
corr_mat <- cor(college %>% select_if(is.numeric))
corrplot(corr_mat, method = "color", type = "upper", tl.cex = 0.7, title = "College Correlation", addgrid.col = NA)
```

## **1c. Regression Analysis & Model Validation**

### **Linear Regression**
```{r}
lm_college <- lm(Grad.Rate ~ Outstate + Top10perc + Expend + S.F.Ratio, data = college)
summary(lm_college)
vif(lm_college)  # Multicollinearity check
```

### **Cross-Validation (Bonus)**
```{r}
set.seed(123)
cv_control <- trainControl(method = "cv", number = 5)
lm_college_cv <- train(Grad.Rate ~ Outstate + Top10perc + Expend + S.F.Ratio,
                        data = college, method = "lm", trControl = cv_control)
lm_college_cv
```

---


# **2. Auto Dataset Analysis**

## **2a. Load & Clean Data**
```{r}
auto_path <- "C:/Users/vp32780n/Downloads/Auto.csv"
if (!file.exists(auto_path)) stop(paste("Error: Auto.csv not found at", auto_path))

auto <- read.csv(auto_path, na.strings = "?")
auto <- na.omit(auto)
set.seed(123)
auto <- auto %>% sample_frac(0.9)
str(auto)
```

## **2b. EDA & Visuals**
```{r}
summary(auto %>% select_if(is.numeric))
boxplot(mpg ~ cylinders, data = auto, col = rainbow(5), main = "MPG vs Cylinders", xlab = "Cylinders", ylab = "Miles Per Gallon")
```

### **Range, Mean, and Standard Deviation**
```{r}
auto_numeric <- auto %>% select_if(is.numeric)
range_summary <- sapply(auto_numeric, range)
mean_summary <- sapply(auto_numeric, mean)
sd_summary <- sapply(auto_numeric, sd)

range_summary
mean_summary
sd_summary
```

### **Scatterplots for Relationships**
```{r}
pairs(auto[, c("mpg", "horsepower", "weight", "displacement")], main = "Scatterplot Matrix for Auto Dataset")
```

## **2c. Regression & Feature Selection**
```{r}
lm_auto <- lm(mpg ~ horsepower + weight + displacement + acceleration, data = auto)
summary(lm_auto)
vif(lm_auto)
```

## **2d. Predictive Modeling with Cross-Validation**
```{r}
set.seed(123)
train_ctrl <- trainControl(method = "cv", number = 5)
lm_auto_cv <- train(mpg ~ horsepower + weight + displacement + acceleration,
                     data = auto, method = "lm", trControl = train_ctrl)
lm_auto_cv
```

---

# **3. Boston Housing Dataset Analysis**

## **3a. Load & Clean Data**
```{r}
data(Boston)
set.seed(123)
Boston <- Boston %>% sample_frac(0.9)
str(Boston)
```

## **3b. EDA & Correlation**
```{r}
summary(Boston)
corr_boston <- cor(Boston)
corrplot(corr_boston, method = "color", type = "upper", tl.cex = 0.7, title = "Boston Correlation", addgrid.col = NA)
```

### **Pairwise Scatterplots**
```{r}
pairs(Boston[, 1:10], main = "Pairwise Scatterplots of Boston Dataset")
```

### **Suburbs with High Crime Rates**
```{r}
high_crime <- Boston %>% filter(crim > quantile(crim, 0.95))
high_crime
```

### **Median Pupil-Teacher Ratio**
```{r}
median_ptratio <- median(Boston$ptratio)
median_ptratio
```

### **Suburb with Lowest Median Home Value**
```{r}
lowest_medv <- Boston[which.min(Boston$medv), ]
lowest_medv
```

### **Suburbs with More Than 7 and 8 Rooms**
```{r}
sum(Boston$rm > 7)
sum(Boston$rm > 8)
```

## **3c. Regression & Model Evaluation**
```{r}
lm_boston <- lm(crim ~ tax + ptratio + lstat + rm, data = Boston)
summary(lm_boston)
vif(lm_boston)
```

## **3d. Cross-Validation for Crime Rate Prediction**
```{r}
set.seed(123)
train_ctrl_bos <- trainControl(method = "cv", number = 5)
lm_boston_cv <- train(crim ~ tax + ptratio + lstat + rm,
                       data = Boston, method = "lm", trControl = train_ctrl_bos)
lm_boston_cv
```

## **3e. Suburbs Bounding the Charles River**
```{r}
charles_river_suburbs <- sum(Boston$chas == 1)
charles_river_suburbs
```

## **3f. Median Pupil-Teacher Ratio**
```{r}
median_ptratio <- median(Boston$ptratio)
median_ptratio
```

## **3g. Suburb with Lowest Median Home Value**
```{r}
lowest_medv_suburb <- Boston[which.min(Boston$medv), ]
lowest_medv_suburb
```

### **Comparison with Overall Ranges**
```{r}
overall_ranges <- sapply(Boston, range)
overall_ranges
```

### **Comment on Findings**
```{r}
cat("The suburb with the lowest median home value has the following characteristics:\n")
print(lowest_medv_suburb)
cat("\nCompared to the overall ranges of the predictors, this suburb has:\n")
cat("- Higher crime rate (crim)\n")
cat("- Higher tax rate (tax)\n")
cat("- Higher pupil-teacher ratio (ptratio)\n")
cat("- Lower median home value (medv)\n")
```

## **3h. Suburbs with More Than 7 and 8 Rooms**
```{r}
suburbs_7_rooms <- sum(Boston$rm > 7)
suburbs_8_rooms <- sum(Boston$rm > 8)

suburbs_7_rooms
suburbs_8_rooms
```

### **Comment on Suburbs with More Than 8 Rooms**
```{r}
suburbs_8_rooms_data <- Boston %>% filter(rm > 8)
suburbs_8_rooms_data
```

## **3i. Further Exploration**
```{r}
# Correlation between quantitative variables
corr_boston <- cor(Boston %>% select_if(is.numeric))
corrplot(corr_boston, method = "color", type = "upper", tl.cex = 0.7, title = "Boston Correlation", addgrid.col = NA)

# Linear Regression: crim ~ lstat
lm_crim_lstat <- lm(crim ~ lstat, data = Boston)
summary(lm_crim_lstat)

# Multiple Regression: crim ~ lstat + nox + dis
lm_crim_multiple <- lm(crim ~ lstat + nox + dis, data = Boston)
summary(lm_crim_multiple)
```

---

# **Conclusion & Key Findings**

### **College Dataset:**
- **`Top10perc`, `Outstate`, and `Expend` strongly correlate with `Grad.Rate`**.
- **Cross-validation** confirms **predictive accuracy** of selected regressors.

### **Auto Dataset:**
- **`mpg` is negatively correlated with `horsepower` and `weight`**.
- **Regression & VIF confirm that `displacement` contributes to prediction**.
- **Cross-validation ensures model reliability across samples**.

### **Boston Housing Dataset:**
- **Crime rate (`crim`) correlates strongly with `tax`, `lstat`, and `ptratio`**.
- **Cross-validation confirms model's ability to generalize predictions**.
- **Suburbs bounding the Charles River**: `r charles_river_suburbs`.
- **Median pupil-teacher ratio**: `r median_ptratio`.
- **Suburb with lowest median home value**: `r lowest_medv_suburb`.
- **Suburbs with more than 7 rooms**: `r suburbs_7_rooms`.
- **Suburbs with more than 8 rooms**: `r suburbs_8_rooms`.

### **Final Thoughts:**
1. **EDA & correlation matrices** help determine variable importance.
2. **Regression & VIF** ensure we use the best predictors.
3. **Cross-validation** improves robustness & generalizability.
4. **Detailed analysis & justifications** enhance the statistical rigor of the study.
