---
title: "Regression Analysis of the College, Auto and Boston Datasets"
subtitle: "Statistical modeling and model evaluation in R"
author: "Vijaysingh Puwar"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: kable
    mathjax: null
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  fig.width = 9,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 96
)

# Locate the project root (the folder that holds R/helpers.R) so the report
# renders the same way from RStudio, from render.R or from any working
# directory.
find_project_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)
  while (!file.exists(file.path(path, "R", "helpers.R"))) {
    parent <- dirname(path)
    if (parent == path) stop("Could not find the project root (R/helpers.R)")
    path <- parent
  }
  path
}
root <- find_project_root()
knitr::opts_knit$set(root.dir = root)
```

```{r load, include = FALSE}
source(file.path(root, "R", "helpers.R"))
source(file.path(root, "R", "analysis.R"))
check_packages()
library(ggplot2)

res <- run_analysis(file.path(root, "data"))
set <- res$settings

fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
pct <- function(x, digits = 0) paste0(fmt(100 * x, digits), "%")
pval <- function(model, term) summary(model)$coefficients[term, "Pr(>|t|)"]
fmt_p <- function(p) if (p < 0.001) "< 0.001" else fmt(p, 3)

show_comparison <- function(comparison) {
  knitr::kable(comparison, digits = 3, row.names = FALSE,
               col.names = c("Model", "CV RMSE (train)", "Test RMSE",
                             "Test MAE", "Test R squared"))
}
coef_table <- function(model, digits = 4) {
  knitr::kable(summary(model)$coefficients, digits = digits)
}
by_model <- function(comparison, model) comparison[comparison$Model == model, ]
best_model <- function(comparison) {
  comparison[which.min(comparison$Test_RMSE), ]
}
gain_vs_baseline <- function(comparison, model) {
  base <- by_model(comparison, "Mean baseline")$Test_RMSE
  1 - by_model(comparison, model)$Test_RMSE / base
}
shapiro_w <- function(model) {
  unname(stats::shapiro.test(rstandard(model))$statistic)
}
n_influential <- function(model) sum(cooks.distance(model) > 4 / nobs(model))

asset_dir <- file.path(root, "assets")
save_asset <- function(plot, name, width = 9, height = 4.5) {
  ggsave(file.path(asset_dir, name), plot, width = width, height = height,
         dpi = 150, bg = "white")
}

col_cmp <- res$college$eval$comparison
auto_cmp <- res$auto$eval$comparison
bos_cmp <- res$boston$eval$comparison

auto_log_name <- "OLS, log(mpg) ~ weight + year + origin"
auto_course_name <- "OLS, 4 coursework predictors"
col_best <- best_model(col_cmp)
auto_best <- best_model(auto_cmp)
bos_best <- best_model(bos_cmp)
bos_ols <- by_model(bos_cmp, "OLS, all 12 predictors")
```

```{r claims-check, include = FALSE}
# Some sentences in this report describe results in words. These checks stop
# the render if a future change to the data or models makes one of them false.
claim <- function(ok, what) {
  if (!isTRUE(ok)) stop("Report claim no longer holds: ", what)
}
p_of <- pval
cd <- res$college$data
claim(nrow(res$college$raw) == 777, "College has 777 rows")
claim(nrow(res$auto$data) == 392, "Auto has 392 rows")
claim(nrow(res$boston$data) == 506, "Boston has 506 rows")
claim(all(tapply(cd$Outstate, cd$Private, median)[["Yes"]] >
            tapply(cd$Outstate, cd$Private, median)[["No"]],
          tapply(cd$Outstate, cd$Elite, median)[["Yes"]] >
            tapply(cd$Outstate, cd$Elite, median)[["No"]]),
      "private and elite colleges have higher median tuition")
claim(coef(res$college$lm_course)[["Expend"]] < 0 &&
        cor(cd$Expend, cd$Grad.Rate) > 0,
      "Expend coefficient negative but simple correlation positive")
claim(p_of(res$college$lm_course, "S.F.Ratio") > 0.05,
      "S.F.Ratio not significant")
claim(max(cooks.distance(res$college$lm_all)) < 0.5,
      "no College observation dominates the fit")
claim(p_of(res$auto$lm_course, "displacement") > 0.05 &&
        p_of(res$auto$lm_course, "acceleration") > 0.05,
      "displacement and acceleration add little to the coursework model")
claim(auto_best$Model == auto_log_name &&
        which.min(auto_cmp$CV_RMSE) == which(auto_cmp$Model == auto_log_name),
      "log(mpg) model is best on both CV and test RMSE")
claim(by_model(auto_cmp, "Lasso, all 7 predictors")$Test_RMSE >
        by_model(auto_cmp, "OLS, weight + year + origin")$Test_RMSE,
      "lasso does not beat the hand-specified models")
claim(by_model(auto_cmp, "OLS, weight + year + origin")$Test_RMSE <
        by_model(auto_cmp, auto_course_name)$Test_RMSE,
      "reduced model improves on the coursework model")
claim(shapiro_w(res$boston$crime_log) > shapiro_w(res$boston$crime_raw),
      "log1p crime residuals closer to normal")
claim(p_of(res$boston$crime_log, "tax") < 0.001 &&
        p_of(res$boston$crime_log, "lstat") < 0.001,
      "tax and lstat strongly associated with log crime")
claim(p_of(res$boston$crime_log, "ptratio") > 0.05 &&
        p_of(res$boston$crime_log, "chas") > 0.05,
      "ptratio and chas not associated with log crime")
hc <- res$boston$data[res$boston$data$crim >
                        quantile(res$boston$data$crim, 0.95), ]
claim(length(unique(hc$tax)) == 1 && length(unique(hc$ptratio)) == 1,
      "high-crime suburbs share one tax rate and one pupil-teacher ratio")
crime_hi <- fitted(res$boston$crime_log) > 1.5
claim(mean(res$boston$data$tax[crime_hi] == 666) > 0.8,
      "high fitted log crime values are mostly the tax = 666 suburbs")
claim(grepl("^Regression tree", bos_best$Model),
      "tree has the lowest Boston test RMSE")
bos_t <- abs(summary(res$boston$lm_all)$coefficients[-1, "t value"])
claim(setequal(names(sort(bos_t, decreasing = TRUE))[1:5],
               c("lstat", "rm", "ptratio", "dis", "nox")),
      "largest Boston OLS t values")
claim(coef(res$boston$lm_all)[["lstat"]] < 0 &&
        coef(res$boston$lm_all)[["ptratio"]] < 0 &&
        coef(res$boston$lm_all)[["rm"]] > 0, "Boston OLS coefficient signs")
```

# Executive summary

This report analyzes three public teaching datasets from *An Introduction to
Statistical Learning* (ISLR) with linear regression, lasso regression and
regression trees. Every model is fitted on an 80% training split, tuned and
compared with 10-fold cross-validation on that split, and scored once on a
held-out 20% test split against a mean-only baseline. All numbers below are
computed when the report is rendered.

- **Auto (fuel efficiency).** The best test RMSE came from
  **`r auto_best$Model`**: `r fmt(auto_best$Test_RMSE)` mpg against
  `r fmt(by_model(auto_cmp, "Mean baseline")$Test_RMSE)` mpg for the
  baseline, a `r pct(gain_vs_baseline(auto_cmp, auto_best$Model))` reduction
  (test R squared `r fmt(auto_best$Test_R2)`). The original four-predictor
  coursework model reached `r fmt(by_model(auto_cmp, auto_course_name)$Test_RMSE)` mpg.
  Displacement, horsepower, weight and cylinders are strongly collinear
  (VIF up to `r fmt(max(res$auto$vif_size), 1)`).
- **Boston (housing values).** The best test RMSE came from
  **`r bos_best$Model`**: `r fmt(bos_best$Test_RMSE)` (in $1000s) against
  `r fmt(by_model(bos_cmp, "Mean baseline")$Test_RMSE)` for the baseline and
  `r fmt(bos_ols$Test_RMSE)` for ordinary least squares (test R squared
  `r fmt(bos_best$Test_R2)`).
- **Boston (crime).** Per capita crime is extremely right skewed. Modeling
  `log1p(crim)` instead of `crim` made the residuals much closer to normal
  (Shapiro-Wilk W `r fmt(shapiro_w(res$boston$crime_raw))` raw vs
  `r fmt(shapiro_w(res$boston$crime_log))` transformed), and property tax rate
  and lower-status population share remained strongly associated with crime.
- **College (graduation rate).** The models explain a moderate share of the
  variation. The best test R squared was `r fmt(col_best$Test_R2)`
  (`r col_best$Model`, test RMSE `r fmt(col_best$Test_RMSE)` percentage
  points against `r fmt(by_model(col_cmp, "Mean baseline")$Test_RMSE)` for the
  baseline), so most variation in graduation rate is not captured by these
  institutional variables.
- **Data quality.** `r nrow(res$college$invalid)` College rows contain
  percentages above 100 and are excluded from modeling, with the raw file left
  unchanged.

These are observational datasets, so every result describes association, not
causation.

# Objectives and questions

| Dataset | Question | Response |
|---|---|---|
| College | Which institutional characteristics are associated with graduation rate, and how well do they predict it? | `Grad.Rate` |
| Auto | Which vehicle characteristics are associated with fuel efficiency, and how does predictor collinearity affect the model? | `mpg` |
| Boston | Which neighborhood characteristics are associated with (a) per capita crime and (b) median home value? | `crim`, `medv` |

The report also keeps the descriptive questions from the original coursework
(for example, how many suburbs bound the Charles River) and answers them once,
with values generated from the data.

# Data

```{r data-summary}
data.frame(
  Dataset = c("College", "Auto", "Boston"),
  File = c("data/College.csv", "data/Auto.csv", "data/Boston.csv"),
  Rows = c(nrow(res$college$raw), nrow(res$auto$data), nrow(res$boston$data)),
  `Rows modeled` = c(nrow(res$college$data), nrow(res$auto$data),
                     nrow(res$boston$data)),
  `Missing values` = c(sum(is.na(res$college$raw)), sum(is.na(res$auto$data)),
                       sum(is.na(res$boston$data))),
  check.names = FALSE
)
```

- **College.** 777 U.S. colleges from the 1995 *US News and World Report*.
  Columns `Top10perc`, `Top25perc`, `PhD`, `Terminal`, `perc.alumni` and
  `Grad.Rate` are percentages and are checked against the range 0 to 100.
- **Auto.** The ISLR `Auto` data. The original file marks five unknown
  horsepower values as `?`; this copy already excludes those rows, and the
  loader still treats `?` as missing so a different copy would fail validation
  instead of being read silently. `origin` is converted from codes 1/2/3 to
  American/European/Japanese.
- **Boston.** The ISLR2 version of the Boston housing data. It is identical to
  `MASS::Boston` except that it omits the `black` column, so the analysis reads
  the shipped CSV and does not load `MASS`.

All loaders stop with an explicit error if a required column is missing,
non-numeric or contains missing values.

# Methodology

- **Split.** Each dataset is split once into 80% training and 20% test rows
  with seed `r set$split_seed`. All models for a dataset use the same split.
- **Cross-validation.** 10-fold cross-validation on the training rows only,
  with one fold assignment (seed `r set$cv_seed`) shared by all models, so CV
  RMSE values are directly comparable. The test rows are used once, for the
  final scores.
- **Models.**
  - *Mean baseline*: predicts the training mean for every row. Any useful
    model has to beat it.
  - *Ordinary least squares (OLS)*: the interpretable reference model,
    including the coursework specifications.
  - *Lasso* (`glmnet`): shrinks coefficients and can set some to zero, which
    helps when predictors are collinear. The penalty is chosen by an inner
    10-fold CV using the one-standard-error rule.
  - *Regression tree* (`rpart`): captures nonlinear relationships and
    interactions without specifying them. It is pruned at the complexity
    parameter with the lowest inner cross-validated error.
- **Metrics.** RMSE and MAE are in the units of the response. Test R squared
  is `1 - SSE/SST` on the test rows; it can be negative when a model is worse
  than predicting the test mean, which is why the baseline scores close to 0.
- **Diagnostics.** Residuals vs fitted values, a normal Q-Q plot and a
  scale-location plot for the main linear models, plus a count of
  observations with Cook's distance above 4/n.

The original coursework randomly removed 10% of each dataset before analysis.
That step is dropped here: evaluation now relies on a held-out test set, and
discarding rows would only reduce the data available for fitting.

# College: graduation rate

## Data checks

```{r college-invalid}
res$college$invalid[, c("Name", college_percent_cols)]
```

These rows report a percentage above 100, which is impossible. They are
excluded from the modeled data (`r nrow(res$college$data)` of
`r nrow(res$college$raw)` rows remain). Because only
`r nrow(res$college$invalid)` rows are affected, excluding them is simpler and
more transparent than guessing a corrected value.

## Exploration

```{r college-eda, fig.height = 4.5}
college_plot <- ggplot(res$college$data,
                       aes(Outstate, Grad.Rate, colour = Private)) +
  geom_point(alpha = 0.55, size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  scale_colour_manual(values = c(No = palette_main[["secondary"]],
                                 Yes = palette_main[["primary"]]),
                      labels = c(No = "Public", Yes = "Private"),
                      name = NULL) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(title = paste("Colleges with higher tuition tend to have higher",
                     "graduation rates"),
       x = "Out-of-state tuition (USD)", y = "Graduation rate (%)",
       caption = "Each point is a college. Lines are least-squares fits.") +
  theme_report()
college_plot
```

```{r college-tuition}
tuition_by_group <- function(group) {
  stats <- tapply(res$college$data$Outstate, res$college$data[[group]],
                  stats::median)
  data.frame(Grouping = group, Level = names(stats),
             `Median out-of-state tuition` = as.numeric(stats),
             check.names = FALSE)
}
rbind(tuition_by_group("Private"), tuition_by_group("Elite"))
```

`Elite` marks colleges where more than 50% of new students came from the top
10% of their high school class (`r sum(res$college$data$Elite == "Yes")`
colleges). Private and elite colleges both have higher median out-of-state
tuition.

```{r college-cor}
col_cor <- stats::cor(res$college$data[college_numeric_cols])[, "Grad.Rate"]
col_cor <- sort(col_cor[names(col_cor) != "Grad.Rate"], decreasing = TRUE)
top_bottom <- c(head(col_cor, 5), tail(col_cor, 5))
knitr::kable(data.frame(Predictor = names(top_bottom),
                        `Correlation with Grad.Rate` = top_bottom,
                        check.names = FALSE),
             digits = 2, row.names = FALSE)
```

The table shows the five strongest positive and five most negative
correlations. On its own, instructional expenditure (`Expend`) is positively
correlated with graduation rate (r = `r fmt(col_cor[["Expend"]])`).

## Coursework model

```{r college-course}
coef_table(res$college$lm_course)
```

Holding the other three predictors fixed, the coefficient on `Expend` is
negative (p = `r fmt_p(pval(res$college$lm_course, "Expend"))`), even though
its simple correlation with graduation rate is positive. `Expend` is
correlated with `Outstate` and `Top10perc`, so its coefficient describes what
remains after those are accounted for, not its overall relationship.
`S.F.Ratio` is not significant in this model
(p = `r fmt_p(pval(res$college$lm_course, "S.F.Ratio"))`).

## Model comparison

```{r college-compare}
show_comparison(col_cmp)
```

All models improve on the baseline, but the gains are modest: the best test R
squared is `r fmt(col_best$Test_R2)`. The VIFs of the full OLS model show
strong collinearity among the size variables (`Accept`
`r fmt(res$college$vif[["Accept"]], 1)`, `Enroll`
`r fmt(res$college$vif[["Enroll"]], 1)`, `F.Undergrad`
`r fmt(res$college$vif[["F.Undergrad"]], 1)`, `Apps`
`r fmt(res$college$vif[["Apps"]], 1)`), which makes their individual
coefficients unstable even when predictions remain reasonable.

## Diagnostics

```{r college-diag}
plot_diagnostics(res$college$lm_all,
                 "Diagnostics: OLS, all 17 predictors (College training set)")
```

Residuals show no clear curvature and are close to normal apart from a few
large positive residuals. Their spread narrows for colleges with high
predicted graduation rates, which is expected for a percentage bounded at 100.
`r n_influential(res$college$lm_all)` training rows have Cook's distance above
4/n, but the largest is `r fmt(max(cooks.distance(res$college$lm_all)))`, so no
single college dominates the fit.

# Auto: fuel efficiency

## Collinearity

```{r auto-cor, fig.height = 5}
plot_correlation(res$auto$data[c("mpg", "cylinders", "displacement",
                                 "horsepower", "weight", "acceleration",
                                 "year")],
                 "Correlations among Auto variables")
```

```{r auto-vif}
knitr::kable(data.frame(
  Predictor = names(res$auto$vif_size),
  `VIF (cylinders, displacement, horsepower, weight)` = res$auto$vif_size,
  check.names = FALSE
), digits = 1, row.names = FALSE)
knitr::kable(data.frame(
  Predictor = names(res$auto$vif_course),
  `VIF (coursework predictors)` = res$auto$vif_course,
  check.names = FALSE
), digits = 1, row.names = FALSE)
```

A variance inflation factor (VIF) measures how much the variance of a
coefficient estimate is inflated because that predictor can be predicted from
the other predictors. It is a multicollinearity diagnostic, not a measure of
importance. Values above roughly 5 to 10 indicate that individual coefficients
are poorly determined. Here the engine-size variables carry largely the same
information.

## Coursework model

```{r auto-course}
coef_table(res$auto$lm_course)
```

With horsepower and weight in the model, displacement
(p = `r fmt_p(pval(res$auto$lm_course, "displacement"))`) and acceleration
(p = `r fmt_p(pval(res$auto$lm_course, "acceleration"))`) add little. That
does not mean they are unrelated to mpg (their simple correlations are
`r fmt(stats::cor(res$auto$data$mpg, res$auto$data$displacement))` and
`r fmt(stats::cor(res$auto$data$mpg, res$auto$data$acceleration))`); it means
their information overlaps with the other predictors.

## A simpler specification

```{r auto-weight, fig.height = 4.5}
auto_plot <- ggplot(res$auto$data, aes(weight, mpg, colour = origin)) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_smooth(aes(group = 1), method = "loess", formula = y ~ x, se = FALSE,
              colour = "grey25", linewidth = 0.9) +
  scale_colour_manual(values = c(American = palette_main[["primary"]],
                                 European = palette_main[["secondary"]],
                                 Japanese = palette_main[["neutral"]]),
                      name = NULL) +
  scale_x_continuous(labels = scales::label_comma()) +
  labs(title = "Fuel efficiency falls as weight rises, along a curve",
       x = "Weight (lb)", y = "Miles per gallon",
       caption = "Grey line: loess smooth over all cars.") +
  theme_report()
auto_plot
```

Because weight already captures most of the engine-size information, a model
with `weight`, model `year` and `origin` is easier to interpret than one with
several collinear size variables. The relationship between weight and mpg is
curved, and the spread of mpg grows with its level, so the same model is also
fitted to `log(mpg)` with predictions converted back to mpg.

```{r auto-diag-reduced}
plot_diagnostics(res$auto$lm_reduced,
                 "Diagnostics: mpg ~ weight + year + origin")
```

```{r auto-diag-log}
plot_diagnostics(res$auto$lm_log,
                 "Diagnostics: log(mpg) ~ weight + year + origin")
```

On the original scale the residuals curve and fan out with the fitted values.
On the log scale the curvature and the change in spread are both reduced.

```{r auto-log-coef}
coef_table(res$auto$lm_log, digits = 5)
```

On the log scale, each additional 1,000 lb is associated with roughly
`r pct(1 - exp(1000 * coef(res$auto$lm_log)[["weight"]]))` lower mpg, and each
later model year with roughly
`r pct(exp(coef(res$auto$lm_log)[["year"]]) - 1, 1)` higher mpg, holding the
other terms fixed.

## Model comparison

```{r auto-compare}
show_comparison(auto_cmp)
```

Dropping redundant size variables and adding `year` and `origin` improves on
the coursework model, and the log specification is the most accurate model on
both CV and test RMSE. The lasso, given all seven predictors, does not beat
the simpler hand-specified models. Removing a variable only because its VIF is
high is not automatically correct; here it is supported by the comparison, but
the reduced model also adds `year`, which accounts for much of the gain.

# Boston: crime and housing values

## Descriptive questions

```{r boston-descriptive}
bos <- res$boston$data
q95 <- stats::quantile(bos$crim, 0.95)
high_crime <- bos[bos$crim > q95, ]
lowest <- bos[which.min(bos$medv), , drop = FALSE]
lowest_pct <- vapply(names(bos), function(v) mean(bos[[v]] <= lowest[[v]]),
                     numeric(1))
over8 <- bos[bos$rm > 8, ]
```

- `r sum(bos$chas == 1)` of `r nrow(bos)` suburbs bound the Charles River.
- The median pupil-teacher ratio is `r fmt(stats::median(bos$ptratio), 1)`.
- `r nrow(high_crime)` suburbs are above the 95th percentile of per capita
  crime (`r fmt(q95)`). All of them have a property tax rate of
  `r paste(sort(unique(high_crime$tax)), collapse = ", ")` per $10,000 and a
  pupil-teacher ratio of
  `r paste(sort(unique(high_crime$ptratio)), collapse = ", ")`.
- `r sum(bos$rm > 7)` suburbs average more than seven rooms per dwelling and
  `r nrow(over8)` average more than eight. The more-than-eight group has a
  median `medv` of `r fmt(stats::median(over8$medv), 1)` versus
  `r fmt(stats::median(bos$medv), 1)` overall, and a median `lstat` of
  `r fmt(stats::median(over8$lstat), 1)`% versus
  `r fmt(stats::median(bos$lstat), 1)`%.

The suburb with the lowest median home value (row `r rownames(lowest)`,
`medv` = `r fmt(lowest$medv, 1)`) is compared with the rest of the data below.
The percentile column is the share of suburbs with a value at or below this
suburb's. The binary `chas` column is left out; this suburb does not bound
the Charles River (`chas` = `r lowest$chas`).

```{r boston-lowest}
cont <- setdiff(names(bos), "chas")
knitr::kable(data.frame(
  Variable = cont,
  `This suburb` = as.numeric(lowest[1, cont]),
  Minimum = vapply(bos[cont], min, numeric(1)),
  Maximum = vapply(bos[cont], max, numeric(1)),
  Percentile = pct(lowest_pct[cont]),
  check.names = FALSE
), digits = 3, row.names = FALSE)
```

## Crime: choosing a scale

```{r boston-crime-dist, fig.height = 3.8}
crime_long <- data.frame(
  scale = factor(rep(c("crim", "log1p(crim)"), each = nrow(bos)),
                 levels = c("crim", "log1p(crim)")),
  value = c(bos$crim, log1p(bos$crim))
)
crime_plot <- ggplot(crime_long, aes(value)) +
  geom_histogram(bins = 35, fill = palette_main[["primary"]],
                 colour = "white") +
  facet_wrap(~scale, scales = "free") +
  labs(title = "Per capita crime is extremely right skewed",
       subtitle = "A log1p transform spreads out the many near-zero values",
       x = NULL, y = "Suburbs",
       caption = paste0("Median crim = ", fmt(stats::median(bos$crim)),
                        ", mean = ", fmt(mean(bos$crim)),
                        ", maximum = ", fmt(max(bos$crim)), ".")) +
  theme_report()
crime_plot
```

The median crime rate is `r fmt(stats::median(bos$crim))` but the mean is
`r fmt(mean(bos$crim))`, pulled up by a small number of extreme suburbs.
`log1p(crim)` (log of 1 + crim) is used because it is defined at zero and
behaves like a log for larger values.

The coursework crime model (`tax`, `ptratio`, `lstat`, `rm`) is fitted to all
506 suburbs on both scales, with `chas` added to test whether bounding the
Charles River is associated with crime once the other variables are included.
These models are for association, not prediction, so they use every row.

```{r boston-crime-diag-raw}
plot_diagnostics(res$boston$crime_raw, "Diagnostics: crim (original scale)")
```

```{r boston-crime-diag-log}
plot_diagnostics(res$boston$crime_log, "Diagnostics: log1p(crim)")
```

On the original scale the residuals are far from normal (Shapiro-Wilk W =
`r fmt(shapiro_w(res$boston$crime_raw))`) and their spread grows sharply with
the fitted values. On the log1p scale the residuals are much closer to normal
(W = `r fmt(shapiro_w(res$boston$crime_log))`) and their spread no longer grows
steadily, so the transformed model is used for interpretation. It is still
not a clean fit: the residuals separate into two groups, and the group with
high fitted values is mostly the high-tax suburbs (tax = 666) discussed
above. A single linear term cannot fully describe that cluster, so the
p-values below should be read as approximate. R squared is
`r fmt(summary(res$boston$crime_raw)$r.squared)` on the original scale and
`r fmt(summary(res$boston$crime_log)$r.squared)` on the log1p scale; because
the responses differ, these two values are not directly comparable.

```{r boston-crime-coef}
coef_table(res$boston$crime_log)
```

Higher property tax rates and a higher share of lower-status population are
associated with higher crime (both p `r fmt_p(pval(res$boston$crime_log, "tax"))`).
Pupil-teacher ratio (p = `r fmt_p(pval(res$boston$crime_log, "ptratio"))`)
and bounding the Charles River (p = `r fmt_p(pval(res$boston$crime_log, "chas"))`)
show no clear association once the other variables are included. Tax rate is
strongly correlated with highway accessibility (`rad`, r =
`r fmt(stats::cor(bos$tax, bos$rad))`), so the tax coefficient partly reflects
that group of high-tax, high-access suburbs.

## Housing values: model comparison

```{r boston-compare}
show_comparison(bos_cmp)
```

```{r boston-medv-diag}
plot_diagnostics(res$boston$lm_all,
                 "Diagnostics: OLS medv ~ all 12 predictors (training set)")
```

The OLS residuals curve at both ends of the fitted range, which suggests
nonlinear relationships (for example with `lstat`) that a linear model cannot
represent. This is consistent with the regression tree scoring
`r fmt(bos_best$Test_RMSE)` against `r fmt(bos_ols$Test_RMSE)` for OLS on the
test set. `medv` is also capped at 50 in the source data
(`r sum(bos$medv == 50)` suburbs), which limits how well any model can fit the
most expensive suburbs.

```{r boston-medv-coef}
coef_table(res$boston$lm_all)
```

In the OLS fit, `lstat`, `rm`, `ptratio`, `dis` and `nox` have the largest t
values. Lower-status share and pupil-teacher ratio are associated with lower
home values and more rooms with higher values, holding the other variables
fixed.

# Model comparison across datasets

```{r overall, fig.height = 5.5}
all_cmp <- rbind(
  cbind(Dataset = "College: Grad.Rate", col_cmp),
  cbind(Dataset = "Auto: mpg", auto_cmp),
  cbind(Dataset = "Boston: medv", bos_cmp)
)
all_cmp$Relative <- ave(all_cmp$Test_RMSE, all_cmp$Dataset,
                        FUN = function(x) x / x[1])
test_sizes <- c(nrow(res$college$eval$test), nrow(res$auto$eval$test),
                nrow(res$boston$eval$test))
plot_cmp <- all_cmp[all_cmp$Model != "Mean baseline", ]
# Order bars within each panel (best at the top); the key keeps model
# names that appear in more than one dataset separate.
plot_cmp$Key <- paste(plot_cmp$Dataset, plot_cmp$Model, sep = "|")
plot_cmp$Key <- stats::reorder(plot_cmp$Key, -plot_cmp$Relative)
plot_cmp$Dataset <- factor(plot_cmp$Dataset,
                           levels = c("Auto: mpg", "Boston: medv",
                                      "College: Grad.Rate"))

comparison_plot <- ggplot(plot_cmp, aes(Relative, Key)) +
  geom_col(fill = palette_main[["primary"]], width = 0.65) +
  geom_text(aes(label = pct(Relative)), hjust = -0.15, size = 3.3) +
  facet_grid(Dataset ~ ., scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^[^|]*\\|", "", x)) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits = c(0, 1.05), expand = c(0, 0)) +
  labs(title = "Test RMSE as a share of the mean-only baseline",
       subtitle = "Lower is better; 100% means no better than the baseline",
       x = "Test RMSE / baseline test RMSE", y = NULL,
       caption = paste0("Each dataset uses one 80/20 split; test sets have ",
                        min(test_sizes), " to ", max(test_sizes), " rows.")) +
  theme_report() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0),
        panel.grid.major.y = element_blank())
comparison_plot
```

```{r save-assets, include = FALSE}
save_asset(comparison_plot, "model-comparison.png", height = 5.5)
save_asset(crime_plot, "boston-crime-distribution.png", height = 3.8)
save_asset(auto_plot, "auto-mpg-weight.png")
```

# Key findings

1. Fuel efficiency is predicted well by a small model: on the test set,
   `r auto_log_name` scored RMSE `r fmt(by_model(auto_cmp, auto_log_name)$Test_RMSE)` mpg
   and R squared `r fmt(by_model(auto_cmp, auto_log_name)$Test_R2)`. The engine
   size variables are collinear, so their individual coefficients should not
   be read as separate effects.
2. For Boston home values, a pruned regression tree had the lowest test RMSE
   among the models tried here (`r bos_best$Model`,
   `r fmt(bos_best$Test_RMSE)`), and OLS residuals show nonlinearity.
3. Boston crime should be modeled on a log scale. On that scale, tax rate and
   lower-status share are strongly associated with crime; Charles River
   proximity is not.
4. Institutional variables explain only part of the variation in college
   graduation rates (best test R squared `r fmt(col_best$Test_R2)`).
5. Simple data checks matter: two impossible percentage values were found and
   excluded, and the Boston CSV was confirmed to match `MASS::Boston` apart
   from the omitted `black` column.

# Limitations

- **Association only.** These are observational data; no result here
  supports a causal claim.
- **One split, small test sets.** Test sets have
  `r nrow(res$auto$eval$test)` (Auto), `r nrow(res$boston$eval$test)` (Boston)
  and `r nrow(res$college$eval$test)` (College) rows, so small differences in
  test RMSE between models are within sampling noise. The CV RMSE column gives
  a second, less noisy view.
- **Dated data.** College is from 1995, Auto from the 1970s and early 1980s,
  and Boston from the 1970 census. Results do not describe current conditions.
- **Censored response.** Boston `medv` is capped at 50.
- **Limited model search.** Only linear, lasso and single-tree models are
  compared; the aim is a transparent comparison, not the most accurate model
  possible.

# Reproducibility

- Raw data are read from `data/` and never modified; cleaning happens in code
  (`R/helpers.R`, `R/analysis.R`).
- Seeds: split `r set$split_seed`, CV folds `r set$cv_seed`, inner tuning
  folds `r set$tune_seed`. Rendering twice gives identical numbers.
- Package versions are pinned in `renv.lock`. Render with `Rscript render.R`
  from the project root.

```{r versions}
pkgs <- c("rmarkdown", "knitr", "ggplot2", "glmnet", "rpart")
data.frame(
  Component = c("R", pkgs),
  Version = c(paste(R.version$major, R.version$minor, sep = "."),
              vapply(pkgs, function(p) as.character(utils::packageVersion(p)),
                     character(1))),
  row.names = NULL
)
```

# References

- James, G., Witten, D., Hastie, T. and Tibshirani, R. *An Introduction to
  Statistical Learning*, 2nd edition. Springer. Datasets from the `ISLR2`
  package.
- Harrison, D. and Rubinfeld, D. L. (1978). Hedonic housing prices and the
  demand for clean air. *Journal of Environmental Economics and Management*,
  5, 81 to 102.
- Friedman, J., Hastie, T. and Tibshirani, R. (2010). Regularization paths for
  generalized linear models via coordinate descent. *Journal of Statistical
  Software*, 33(1).
- Therneau, T. and Atkinson, B. `rpart`: Recursive Partitioning and
  Regression Trees. R package.

This project began as a midterm project for CS601C Computational Statistics
at Pace University (Spring 2025) and was later extended with the evaluation,
diagnostics and reproducibility work described above.
